Small induced universal graphs

### 2021-08-24 paper/paper.tex
Get upper and lower bounds for f(6)

\documentclass[12pt]{article}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage[x11names, svgnames, rgb]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{snakes,arrows,shapes}
\usepackage{amsmath}
%

@@

\maketitle
%
\section{Introduction}

Given a collection $\calF$ of graphs, graph $G$ is \emph{induced
universal} for $\calF$ if and only if each element of $\calF$ is an induced
subgraph of $G$.  Graph $G$ is a \emph{minimal} induced universal graph
for $\calF$ is it has as few vertices as possible.  We write $f(k)$ to denote
    the order (number of vertices) of a minimal induced universal graph for the
    family of all graphs on $k$ vertices.  Moon showed that $f(k) \leq 2^{(k-1)/2}$
    \cite{moon_1965}, and Alon showed that $f(k) = (1 + o(1))2^{(k-1)/2}$
    \cite{alon2017asymptotically}.  There is an extensive literature on bounds
    on the order of minimal induced universal subgraphs for many families of
    graph; see the references in \cite{alon2017asymptotically}.  However, I am
    not aware of any existing systematic attempts to find exact values for
    families of small graphs.  This paper presents exact values of $f(k)$ for
    $k \leq 5$, obtained by brute-force computational search.  The
    corresponding counts of minimal induced universal subgraphs, and results for
    families of trees, are also
    presented.

@@

A simple brute-force approach was used to find the set of
$n$-vertex graphs that are induced universal for the family $\calF$ of all $k$-vertex
graphs.  Each possible $n$-vertex graph $G$ from Brendan McKay's lists of
graphs\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/graphs.html}} was
tested in turn.  We used a Python implementation of the McSplit
algorithm \cite{DBLP:conf/ijcai/McCreeshPT17} to test whether each $k$-vertex
graph is an induced subgraph of $G$.  Although the McSplit algorithm was designed for
the more general maximum common induced subgraph problem, it performs well on
the subgraph isomorphism problem if both input graphs are small.

@@

The full set of experiments described in this paper, run sequentially, required
around 14 million calls to the subgraph isomorphism algorithm and took around
12 minutes to complete on a laptop.

@@

\caption{Induced universal graphs for the families of all

@@

>>>>

\documentclass[12pt]{article}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage[x11names, svgnames, rgb]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{amsthm}
\usetikzlibrary{snakes,arrows,shapes}
\usepackage{amsmath}
%

@@

\newtheorem{proposition}{Proposition}

\maketitle
%
\begin{abstract}
Using the results of brute-force computational experiments, we give for $0 \leq
k \leq 5$ the minimum number of vertices $f(k)$ in a graph containing all
$k$-vertex graphs as induced subgraphs.  We give the corresponding results for
$0 \leq k \leq 6$ for graphs containing all trees on $k$ vertices.  We also show
that $13 \leq f(6) \leq 15$.
\end{abstract}

\section{Introduction}

Given a collection $\calF$ of graphs, graph $G$ is \emph{induced
universal} for $\calF$ if and only if each element of $\calF$ is an induced
subgraph of $G$.  Graph $G$ is a \emph{minimal} induced universal graph
for $\calF$ if it has as few vertices as possible.  We write $f(k)$ to denote
    the order (number of vertices) of a minimal induced universal graph for the
    family of all graphs on $k$ vertices.  Moon showed that $f(k) \leq 2^{(k-1)/2}$
    \cite{moon_1965}, and Alon showed that $f(k) = (1 + o(1))2^{(k-1)/2}$
    \cite{alon2017asymptotically}.  There is an extensive literature on bounds
    on the order of minimal induced universal subgraphs for many families of
    graphs; see the references in \cite{alon2017asymptotically}.  However, I am
    not aware of any existing systematic attempts to find exact values for
    families of small graphs.  This paper presents exact values of $f(k)$ for
    $k \leq 5$, obtained by brute-force computational search.  The
    corresponding counts of minimal induced universal subgraphs, and results for
    families of trees, are also
    presented.

@@

Given values of $k$ and $n$,
a simple brute-force approach was used to find the set of
$n$-vertex graphs that are induced universal for the family $\calF$ of all $k$-vertex
graphs.  Each possible $n$-vertex graph $G$ from Brendan McKay's lists of
graphs\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/graphs.html}} was
tested in turn.  We used a Python implementation of the McSplit
algorithm \cite{DBLP:conf/ijcai/McCreeshPT17} to test whether each $k$-vertex
graph is an induced subgraph of $G$.  Although the McSplit algorithm was designed for
the more general maximum common induced subgraph problem, it performs well on
the subgraph isomorphism problem if both input graphs are small.

@@

The full set of experiments described in this paper, run sequentially, required
around 14 million calls to the subgraph isomorphism algorithm and took around
12 minutes to complete on a laptop.  \footnote{The code and results from this
paper, including lists of minimal induced universal graphs graph6 format, are
available from \url{https://github.com/jamestrimble/small-universal-graphs}}.

@@

\caption{Minimal induced universal graphs for the families of all

@@

\section{Bounds on $f(6)$}

To give an upper bound of 15 on $f(6)$, Figure~\ref{fig:adjmat15} shows the
adjacency matrix of a 15-vertex graph that is induced universal for the family
of all graphs on six vertices.  This was generated using a specialised random
graph generator.  We number the vertices from zero; the generator makes the
$k$ vertices numbered $0$ to $k-1$ a clique, and the $k$ vertices numbered $k-1$ to $2k-2$ an
independent set.  Each possible edge that is not involved in either the clique
or the independent set is added with probability $1/2$.  It was only necessary
to generate 53 graphs with this generator to find a solution.

\begin{figure}[h]
\centering
\verb|0 1 1 1 1 1 0 0 0 1 1 0 0 0 1| \\
\verb|1 0 1 1 1 1 0 0 0 0 1 1 1 1 1| \\
\verb|1 1 0 1 1 1 1 0 0 1 0 0 1 0 0| \\
\verb|1 1 1 0 1 1 1 1 1 1 0 0 0 0 0| \\
\verb|1 1 1 1 0 1 1 1 0 1 1 1 1 0 0| \\
\verb|1 1 1 1 1 0 0 0 0 0 0 0 0 0 1| \\
\verb|0 0 1 1 1 0 0 0 0 0 0 0 0 1 0| \\
\verb|0 0 0 1 1 0 0 0 0 0 0 1 1 1 1| \\
\verb|0 0 0 1 0 0 0 0 0 0 0 1 1 1 0| \\
\verb|1 0 1 1 1 0 0 0 0 0 0 1 0 0 0| \\
\verb|1 1 0 0 1 0 0 0 0 0 0 0 1 0 1| \\
\verb|0 1 0 0 1 0 0 1 1 1 0 0 0 1 0| \\
\verb|0 1 1 0 1 0 0 1 1 0 1 0 0 1 1| \\
\verb|0 1 0 0 0 0 1 1 1 0 0 1 1 0 0| \\
\verb|1 1 0 0 0 1 0 1 0 0 1 0 1 0 0|
\caption{The adjacency matrix of a 15-vertex induced universal graph for the family of all
six-vertex graphs}
\label{fig:adjmat15}
\end{figure}

For $k > 5$, it is straightforward to increase by 1 the lower bound by ``bof'' from the previous
section.

\begin{proposition}
    $f(k) > 2k + 1$ for all $k > 5$.
\end{proposition}
\begin{proof}
    Suppose, for contradition, that $G$ is an induced universal graph on $2k$
    vertices for the family of all graphs on $k$ vertices.  Graph $G$ must have
    $K_k$ and $I_k$ as induced sugraphs.  This clique and independent set must
    overlap by exactly one vertex (otherwise, as in the proof by ``bof'' from
    the previous section, $G$ would be a split graph and would not have $C_4$
    as an induced subgraph).   Thus we can partition the vertices of $G$ into
    three sets: a clique $S_1$, an independent set $S_2$, and a set $S_3$
    containing the single remaining vertex.

    Let $S$ be a set of vertices such that the subgraph of $G$ induced by $S$
    is isomorphic to the cycle graph $C_k$.  $S$ may contain at most two vertices
    from $S_1$, and these must be adjacent in the induced cycle.  Trivially,
    $S$ can contain at most one vertex from $S_3$.  Hence, $S$ must contain at
    least $k-3$ vertices from $S_2$.  At least two of these vertices must be adjacent
    in the induced $C_k$.  Since $S_2$ is an independent set in $G$, we have a contradiction.
\end{proof}

Thus, we have that $13 \leq f(6) \leq 15$.
